---
title: 'compgen2021: Week 2 exercises'
author: 'ENTER YOUR NAME HERE'
output:
  pdf_document: default
  pdf: default
---

# Exercises for Week2

For this set of exercises we will be using the expression data shown below:
```{r dataLoadClu,eval=FALSE}
expFile=system.file("extdata",
                    "leukemiaExpressionSubset.rds",
                    package="compGenomRData")
mat=readRDS(expFile)

```

### Clustering

1. We want to observe the effect of data transformation in this exercise. Scale the expression matrix with the `scale()` function. In addition, try taking the logarithm of the data with the `log2()` function prior to scaling. Make box plots of the unscaled and scaled data sets using the `boxplot()` function. [Difficulty: **Beginner/Intermediate**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


2. For the same problem above using the unscaled data and different data transformation strategies, use the `ward.d` distance in hierarchical clustering and plot multiple heatmaps. You can try to use the `pheatmap` library or any other library that can plot a heatmap with a dendrogram. Which data-scaling strategy provides more homogeneous clusters with respect to disease types? [Difficulty: **Beginner/Intermediate**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


3. For the transformed and untransformed data sets used in the exercise above, use the silhouette for deciding number of clusters using hierarchical clustering. [Difficulty: **Intermediate/Advanced**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


4. Now, use the Gap Statistic for deciding the number of clusters in hierarchical clustering. Is the same number of clusters identified by two methods? Is it similar to the number of clusters obtained using the k-means algorithm in the unsupervised learning chapter. [Difficulty: **Intermediate/Advanced**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


### Dimension reduction
We will be using the leukemia expression data set again. You can use it as shown in the clustering exercises.

1. Do PCA on the expression matrix using the `princomp()` function and then use the `screeplot()` function to visualize the explained variation by eigenvectors. How many top components explain 95% of the variation? [Difficulty: **Beginner**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


2. Our next tasks are removing the eigenvectors and reconstructing the matrix using SVD, then we need to calculate the reconstruction error as the difference between the original and the reconstructed matrix. HINT: You have to use the `svd()` function and equalize eigenvalue to $0$ for the component you want to remove. [Difficulty: **Intermediate/Advanced**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


3. Produce a 10-component ICA from the expression data set. Remove each component and measure the reconstruction error without that component. Rank the components by decreasing reconstruction-error. [Difficulty: **Advanced**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


4. In this exercise we use the `Rtsne()` function on the leukemia expression data set. Try to increase and decrease perplexity t-sne, and describe the observed changes in 2D plots. [Difficulty: **Beginner**]

**solution:**
put your text here
```{r,echo=TRUE,eval=TRUE}
# FILL IN HERE
 
```


